Estimating cardiac contraction through high resolution data assimilation of a personalized mechanical model

### 2017-07-20 chapters/paper2/main.tex
Add start of acknowledgement and list of papers

% Template for Elsevier CRC journal article
% version 1.1 dated 16 March 2010

% This file (c) 2010 Elsevier Ltd.  Modifications may be freely made,
% provided the edited file is saved under a different name

% This file contains modifications for Procedia Computer Science
% but may easily be adapted to other journals

% Changes since version 1.0
% - elsarticle class option changed from 1p to 3p (to better reflect CRC layout)

%-----------------------------------------------------------------------------------

%% This template uses the elsarticle.cls document class and the extension package ecrc.sty
%% For full documentation on usage of elsarticle.cls, consult the documentation "elsdoc.pdf"
%% Further resources available at http://www.elsevier.com/latex

%-----------------------------------------------------------------------------------

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Important note on usage                  %%
%% -----------------------                  %%
%% This file must be compiled with PDFLaTeX %%
%% Using standard LaTeX will not work!      %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% The '3p' and 'times' class options of elsarticle are used for Elsevier CRC
% \documentclass[review]{elsarticle}
\documentclass[times]{elsarticle}

%% Give the name of the journal
\journal{Journal of Computational Science}

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
\usepackage{amsmath}

% Path to the figures
\graphicspath{{figures/}}

% if you have landscape tables
% \usepackage[figuresright]{rotating}
\usepackage{color}

% Mathetmatical symbols
\newcommand{\tr}{\mathrm{tr} \hspace{0.05cm} }
\newcommand{\argmax}{\mathrm{argmax} \hspace{0.05cm} }

% End-systolic elastance
% \newcommand{\Es{E_{\text{ES}}}}
% \newcommand{\es}{_{\max}}
\newcommand{\es}{_{\text{ES}}}

% Mechanics
\newcommand{\Xvec}{\mathbf{X}}
\newcommand{\xvec}{\mathbf{x}}
\newcommand{\uvec}{\mathbf{u}}
\newcommand{\N}{\mathbf{N}}
\newcommand{\plv}{p_{\textnormal{\tiny{lv}}}}

% Functionals
\newcommand{\Istrainavg}{\overline{I}_{\text{strain}}}
\newcommand{\Istrainrelmax}{\overline{I}^{\text{relmax}}_{\text{strain}}}
\newcommand{\Ivolavg}{\overline{I}_{\text{vol}}}

% Fiber -sheets
\newcommand{\ef}{\mathbf{f}_0}
% \newcommand{\es}{\mathbf{s}_0}
% \newcommand{\en}{\mathbf{e}_n}
\newcommand{\Fef}{\mathbf{f}}

% Local basis
\newcommand{\ec}{\mathbf{e}_c}
\newcommand{\V}[1]{\bm{#1}}

% Tensors
\newcommand{\C}{\mathbf{C}}
\newcommand{\F}{\mathbf{F}}
\newcommand{\I}{\mathbf{I}}
\newcommand{\T}{\mathbf{T}}
\newcommand{\B}{\mathbf{B}}

% Stress tensors
\newcommand{\FPiola}{\mathbf{P}}
\newcommand{\SPiola}{\mathbf{S}}
\newcommand{\Cauchy}{\mathbf{\sigma}}

\begin{document}

\begin{frontmatter}

\title{Estimating cardiac contraction through high resolution data
  assimilation of a personalized mechanical model}

\author[simula,uioinf,cci]{Henrik Finsberg\corref{cor1}\fnref{label1}}
\ead{henriknf@simula.no}
\author[kcl]{Gabriel Balaban}
\author[uiomed,cci]{Hans Henrik Odland}
\author[uiomed,cci]{Stian Ross}
\author[simula,uioinf,cci]{Joakim Sundnes}
\author[simula,aas,cci]{Samuel Wall}

\address[simula]{Simula Research Laboratory, P.O. Box 134 1325
  Lysaker, Norway}
\address[uioinf]{Department of Informatics, University of Oslo,
  P.O. Box 1080, Blindern 0316 Oslo, Norway}
\address[uiomed]{Faculty of Medicine, University of Oslo, P.O. Box
  1078 Blindern, 0316 Oslo, Norway}
\address[cci]{Center for Cardiological Innovation, Songsvannsveien 9,
  0372 Oslo, Norway}
\address[aas]{Department of Mathematical Science and Technology,
  Norwegian University of Life Sciences,  Universitetstunet 3 1430 Ås,
  Norway}
\address[kcl]{Department of Imaging Sciences and Bioengineering,
              King's College London, St. Thomas Hospital, 
              Westminster Bridge Rd, Lambeth, London SE1 7EH, UK}

\cortext[cor1]{Corresponding author}

\begin{abstract}

Cardiac computational models, individually personalized, can provide
clinicians with useful diagnostic information and aid in
treatment planning.  A major bottleneck in this process can be  
determining model parameters to fit created models to individual
patient data. However, adjoint-based data assimilation techniques can
now rapidly estimate high dimensional parameter sets.  This method is
used on a cohort of heart failure patients, capturing cardiac mechanical
information and comparing it with a healthy control group.  Excellent
fit ($R^2 \geq 0.98$) to systolic strains is obtained, and analysis
shows a significant difference in estimated contractility between the
two groups.

\end{abstract}

\begin{keyword}
%% keywords here, in the form: keyword \sep keyword

%% MSC codes here, in the form: \MSC code \sep code
%% or \MSC[2008] code \sep code (2000 is the default)
Cardiac Mechanics \sep Adjoint Method \sep Data assimilation \sep
PDE-constrained optimization \sep Contractility 
\end{keyword}

\end{frontmatter}

% \pagebreak
%%
%% Start line numbering here if you want

% \linenumbers

%% main text
\section{Introduction}
\label{sec:introduction}

@@

Clinical measurements of cardiac function for seven LBBB patients were
obtained from the Impact study \cite{ImpactStudy2016}.
Data was also acquired for seven healthy volunteers. 4D echocardiographic
images of the left ventricle (LV), for both the LBBB patients and
healthy subjects, were captured using
a GE Vingmed E9 device, and analysis carried out with the
software package EchoPac. For each subject, depending on frame rate and cardiac cycle time, the
analysis provided between 15 and 50 LV volumes,  geometric segmentations of the LV
endocardium and epicardium, and cardiac strain calculated via speckle
tracking. The strains were defined according to
the 17 segment AHA-zone representation
\cite{cerqueira2002standardized}, in the longitudinal, radial and
circumferential direction, giving a total of 51 strain measurements
per time point.

The LBBB patients had LV pressure measurements taken during
implantation of a cardiac resynchronization therapy (CRT) device, and
valvular events were used to synchronize the pressure to the echo data
with the help of a medical doctor.  Meanwhile, in the 
healthy control group, where invasive pressure measurements were
absent, the pressure waveform from one of the LBBB patients was used and
scaled to reported values of the end-diastolic and end-systolic
left ventricular pressure [Table 30-1 in \cite{klingensmith2008washington}].

\subsection{Automated geometry and microstructure creation}
For each patient, a 3D tetrahedral mesh of the LV was
constructed using Gmsh \cite{geuzaine2009gmsh} from triangulated
segmented surfaces of the endo- and epicardium at the beginning of
atrial systole, Figure \ref{fig:pipeline}. A cut was made at the
ventricular base of the segmentation, so that the mesh cavity volume
and the ultrasound measured volume differed by less than  $1$ ml. Mesh
cells were marked into the 17 AHA regions through the regionally
delineated strain data, and the myocardial fiber orientation were
assigned using the algorithm from Bayer et al \cite{bayer2012novel},
with the endo- and epicardial helix fiber angles set to
$\alpha_{\text{endo}} = 60$ and $\alpha_{\text{epi}} = -60$, respectively.

\subsection{Mechanical Model}
We represent the heart as a hyperelastic continuum body, where the coordinates in
the reference($\Xvec$) and the current($\xvec$) configuration are
related via the displacement field $\uvec = \xvec-\Xvec$. Furthermore,
we utilize the
deformation gradient, the determinant of the deformation gradient
and, the right Cauchy-Green deformation tensor given by $\F= \I+\nabla
\uvec $,  $J = \det \F$ and $\C = \F^T\F$, respectively. 
% Constitutive relations
To model the passive behavior of the myocardium, the transversely 
isotropic strain energy function proposed in  \cite{holzapfel2009constitutive} is adopted:
\begin{align}
\label{eq:hoa}
  \mathcal{W}(I_1, I_{4\ef}) = \frac{a}{2 b} \left( e^{ b (I_1 
  - 3)}  -1 \right)
  + \frac{a_f}{2 b_f} \left( e^{ b_f (I_{4\ef} 
  - 1)_+^2} -1 \right).
\end{align}
Here $I_1 = \tr \C$ and $I_{4\ef}= \ef \cdot (\C \ef)$ are invariants
of $\C$, $(\cdotp)_{+}  = \max\{\cdot, 0\}$, and $a, a_f, b, b_f$ are
material stiffness parameters defining the elastic properties of the myocardium.
% Incompressibility
We follow a common approach and assume that the myocardium is incompressible. 
Incompressibility is incorporated in the model by using a two-field variational 
approach, where we introduce a Lagrange multiplier $p$ which represents the 
hydrostatic pressure, and the term $p(J-1)$ is added to the strain-energy.
% Active response
To model the active response we apply the active strain framework
\cite{ambrosi2011electromechanical}, decomposing the deformation into
elastic and active components, $\F = \F_e \F_a$. The active part,
$\F_a$ represents the actual shortening along the muscle fibers, while
the elastic part $\F_e$ ensures compatibility of the tissue. We adopt
the formulation of the active deformation gradient from
\cite{gjerald2014patient}, with the following form:  
\begin{equation}
  \F_a =
  (1 - \gamma) \ef \otimes \ef
  + \frac{1}{\sqrt{1 - \gamma}} (\I - \ef \otimes \ef).
 \label{eq:active_strain_Fa_gjerald}
\end{equation}
%which accounts for the effect of shortening along muscle fibers \cite{gjerald2014patient},
%while the elastic part $\F_e$ captures the elastic deformation of the tissue. 
In this formulation of $\F_a$, $\gamma$ prescribes the relative active shortening
along the fibers, and is implemented as a piecewise linear function over
the domain, yielding one parameter per vertex in the mesh. The active
model is coupled to the passive model by letting
$\mathcal{W}$ in \eqref{eq:hoa} be a function of the
modified elastic invariants $I_1^E = \tr \C_e$ and $I_{4\ef}^E = \ef
\cdot (\C_e \ef)$, where $\C_e = \F_e^T\F_e$.

The resulting displacement field $\uvec$ and hydrostatic pressure $p$
are determined by using the principle of stationary potential energy
\cite{holzapfel2000nonlinear}, which is based on minimizing the total
energy  $\Pi(\uvec, p)$, which includes internal energy derived from
\eqref{eq:hoa} and external energy. The external energy includes
contributions from the measured cavity pressure $p_{\text{LV}}$, and a
linear spring term at the basal boundary, with spring constant
$k$ = 1 kPa. The equilibrium solution is found by solving for the minimum
potential energy, $\delta\Pi(\uvec, p) = 0$.

@@

Here $V$ and $\varepsilon_{kj}$ are the measured volume and regional Lagrangian strain in
segment $j$ in direction $k$ respectively, and $\tilde{V}^i =
-\frac{1}{3} \int_{\partial \Omega_{\text{endo}}} (\Xvec + \uvec)
\cdotp J\F^{-T}\N  \mathrm{d}S$,  and $\varepsilon_{kj} =
\frac{1}{|\Omega_j|}\int_{\Omega_j}  \mathbf{e}_k^T \nabla \uvec \,
\mathbf{e}_k  \mathrm{d}x.$ 
The parameters $\alpha$ and $\lambda$
control the weights on the different terms, and the sum in the second
term is taken over the seventeen AHA
segments, and the three different strain components (Section \ref{sec:clinical_data}).

The data assimilation procedure is divided into two phases; a passive
phase and an active phase. For the passive phase we find the stiffness
of the overall ventricle,  using the control parameter $m = a$, where
$a$ is the linear isotropic parameter in \eqref{eq:hoa}, $\alpha=
1.0$,  with $\lambda = 0$ and $\gamma = 0$, minimizing only the misfit
with the measured volumes. The remaining material parameters are fixed
according to [Table 1 row 3 of \cite{holzapfel2009constitutive}].
For the active phase we fix the material
parameter optimized in the passive phase, and the control parameter
becomes the pointwise relative mechanical contraction, $m = \gamma$,
with $\alpha = 0.95$ and $\lambda = 0.01$. This choice of $\alpha$ and
$\lambda$ was based on the analysis done in \cite{balaban}. A summary
of our optimization pipeline is provided to the right in
Figure~\ref{fig:pipeline}.

\begin{figure}[htbp]
\centering
    \includegraphics[width=\textwidth]{models}
\caption{Left: Automated anatomical modeling pipeline to produce AHA
  marked simulation meshes with applied fiber orientations from 3D
  echocardiographic segmentations. Right: Optimization
  pipeline. 1. Optimization of passive model is
  initialized. 2. Linear isotropic material parameter
  $a$ in \eqref{eq:hoa} as control parameter, minimize the misfit
  between simulated and measured volume from beginning of atrial
  systole to end diastole. 3. Fix the optimized value of $a$, let
  $\gamma$ from \eqref{eq:active_strain_Fa_gjerald} be zero,
  and initialize the active model. 4. Using $\gamma$ from
  the previous point as initial guess, minimize the misfit between
  volume and regional strain using $\gamma$ as control. Continue until
  all points in the cardiac cycle are optimized.}
\label{fig:pipeline}
\end{figure}

\subsection{Finite element solution}
We employ a Galerkin finite element method with Taylor-Hood
tetrahedral elements, that is $(\uvec, p) \in \mathbb{P}_2 \times
\mathbb{P}_1 $, with $\mathbb{P}_n$ being the space of piecewise
polynomials of degree $n$.
The solver is implemented in the finite element framework FEniCS
\cite{logg2012automated}, and uses a Newton trust region algorithm
\cite{PETScPackage} to solve nonlinear systems. The minimization of the
model-data misfit functional~\eqref{eq:pde_opt} is accomplished by a sequential
quadratic programming algorithm (SQP)~\cite{kraft1988software}, where
the functional gradient is computed by solving an automatically derived adjoint
equation \cite{farrell2013automated}. In these minimizations an upper bound of $0.4$ is
set for the control variable $\gamma$.

\subsection{Contraction analysis}
The optimized spatially resolved contraction parameter $\gamma$ from 
\eqref{eq:active_strain_Fa_gjerald}, may be interpreted as an
index of contractility, as it describes how much the material attempts
to contract independent of elastic load.  Since this parameter is
defined at all the nodal points in the geometry, we summarize the
contraction of the LV by averaging over the entire ventricular 
domain, $ \overline{\gamma}(t) = \frac{1}{|\Omega|} \int_{\Omega}
\gamma (t, \mathbf{x}) \mathrm{d}\mathbf{x}$, at each time point 
to provide a trajectory of average ventricular contraction for
each patient. In addition, the overall elastic state of the optimized patient models
can be used to give estimates of LV elastance. The left ventricular
end-systolic elastance $E\es$, the response of end systolic volume to
increased load, is considered to be one of the major determinants for
cardiac systolic function, and was in \cite{sagawa1977end} proposed
as a global index of ventricular contractility. 
It is possible to calculate the end systolic elastance directly 
if the end systolic pressure is known or estimated, 
by perturbing the loading conditions on the optimized model at end
systole, and calculating the slope in the resulting
ES pressure-volume curve. More precisely, if $\plv^{\text{ES}}$ is the
end-systolic ventricular pressure, with cavity volume $V^{\text{ES}}$,
we change the pressure to $\plv^{\text{ES} + \Delta} =
\plv^{\text{ES}} + \Delta \plv$, resulting in a change in volume,
$V^{\text{ES} +\Delta} = V^{\text{ES}} + \Delta V$. The end-systolic
elastance can then be estimated by 
\begin{align}
  \tilde{E}\es = \frac{\Delta \plv}{\Delta V}.
  \label{eq:estimate_elastance}
\end{align}

@@

\subsection{Matching of strain and volume}
We show the results from two representative simulations in
Figure~\ref{fig:generic_result}, one from the LBBB group and one from
the healthy control group. Snapshots from the calculated end-diastolic and
end-systolic configurations are depicted, and agreement with PV-loops
and one of the 51 strain traces over the cardiac cycle shown.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\textwidth]{generic_result}
  \caption{Left: Snap shots of end-diastolic(ED) and end-systolic(ES)
    model configurations for an LBBB patient and a healthy
    control. The colormaps shows longitudinal strain. Right: Simulated
    and measured or estimated pressure-volume loops for these hearts,
    as well as the longitudinal strain traces for both in the mid
    septal segment, one of the 51 strain traces optimized. This mid
    septal segment is highlighted in green in the left figure.}
  \label{fig:generic_result}
\end{figure}

The total analysis of the 14 patients involved optimizing 432 volume
measurements and 20 853 strain measurements. The average time for one
forward and gradient evaluation was 8.3 and 8.9 seconds respectively
when running on a cluster using four cores. Here the average number of
control parameters was 985.

In order to visualize the overall match of simulated to measured data,
we show linear regression plots in Figure~\ref{fig:data_matching}. For
the strain, we separately consider the diastolic and systolic points,
as different types of data were used to  constrain the model in these
two phases, namely volume in the diastolic phase and strain in the
systolic phase. An excellent overall fit was obtained for the optimized
volume ($R^2=1.00$) and systolic strains ($R^2=0.98$). 
Diastolic strains, not used in the optimization, were less well
matched ($R^2=0.29$).

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\textwidth]{data_matching}
  \caption{Left: Scatter plot of simulated versus
    measured volumes and the best linear least-squares fit of these
    points, (slope = 1.00). Right: Scatter plot of simulated versus
    measured strain for all segments and all directions, separated
    into the diastole, were only the volume was optimized and systole,
    where both the strain and volume were optimized. The best linear
    least-squares fit for the two cases. For diastole, the slope of
    the best linear fit was $0.35$, while the linear fit for the
    systolic points had a slope of $0.98$.}
  \label{fig:data_matching}
\end{figure}

@@

Global contraction time courses, $\overline{\gamma}$, for each 
patient were synchronized to the valvular
events to normalize for differing cycle lengths. The average 
and standard deviation of these normalized traces for the
LBBB vs the healthy controls are shown in right of
Figure~\ref{fig:contractility}. The healthy patients had a much higher
level of contraction through the cardiac cycle, and the peak values
were compared using one-way ANOVA, yielding a $P-$value less than
$0.001$.

The values of calculated $\tilde{E}\es$ for the healthy and LBBB patients are
shown to the left in Figure~\ref{fig:contractility}. The calculated
elastances of the LBBB group were significantly lower than for the healthy
group, with the comparison between the groups using one-way ANOVA
giving a $P-$value of $0.047$.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\textwidth]{contractility}
  \caption{Left: Estimated values of $\tilde{E}\es$, given by
    \eqref{eq:estimate_elastance}. The mean value is depicted for each
    group as a bar. Right: Mean value of $\gamma$ for the two groups
    synchronized with respect to valvular events (mvc: mitrial valve
    closure, avo: aortic valve opening, avc: aortic valve closure,
    mvo: mitrial valve opening). Shaded area represents $\pm 1$
    standard deviation.} 
  \label{fig:contractility}
\end{figure}

>>>>

 
\graphicspath{{chapters/paper2/figures/}}

\chapter{Paper 2: \\Estimating cardiac contraction through high resolution data
  assimilation of a personalized mechanical model}

\clearpage

\section*{Estimating cardiac contraction through high resolution data
  assimilation of a personalized mechanical model}

\begin{center}
  Henrik Finsberg
  Gabriel Balaban,
  Stian Ross,
  Trine H\r{a}land,
  Hans Henrik Odland, 
  Joakim Sundnes, and
  Samuel Wall
\end{center}

\subsection*{Abstract}
% \begin{abstract}

Cardiac computational models, individually personalized, can provide
clinicians with useful diagnostic information and aid in
treatment planning.  A major bottleneck in this process can be  
determining model parameters to fit created models to individual
patient data. However, adjoint-based data assimilation techniques can
now rapidly estimate high dimensional parameter sets.  This method is
used on a cohort of heart failure patients, capturing cardiac mechanical
information and comparing it with a healthy control group.  Excellent
fit ($R^2 \geq 0.95$) to systolic strains is obtained, and analysis
shows a significant difference in estimated contractility between the
two groups.

\clearpage
%% main text
\section{Introduction}
\label{sec:introduction}

@@

Clinical measurements of cardiac function for seven LBBB patients were
obtained from the Impact study \cite{ImpactStudy2016}.
Data was also acquired for seven healthy volunteers. 4D echocardiographic
images of the left ventricle (LV), for both the LBBB patients and
healthy subjects, were captured using
a GE Vingmed E9 device, and analysis carried out with the
software package EchoPac. For each subject, depending on frame rate and cardiac cycle time, the
analysis provided between 15 and 50 LV volumes,  geometric segmentations of the LV
endocardium and epicardium, and cardiac strain calculated via speckle
tracking. The strain were defined according to
the 17 segment AHA-zone representation
\cite{cerqueira2002standardized}, in the longitudinal, radial and
circumferential direction, giving a total of 51 strain measurements
per time point, with the reference time point for strain analysis
  being the first frame after onset of QRS.

The LBBB patients had LV pressure measurements taken during
implantation of a cardiac resynchronization therapy (CRT) device, and
valvular events were used to synchronize the pressure to the echo
data. In the healthy control group, where invasive pressure measurements were
absent, the pressure waveform from one of the LBBB patients was used and
scaled to reported values of the end-diastolic and end-systolic
left ventricular pressure [Table 30-1 in \cite{klingensmith2008washington}].

\subsection{Automated geometry and microstructure creation}
For each patient, a 3D tetrahedral mesh of the LV was
constructed from triangulated segmented surfaces of the endo- and
epicardium corresponding to the frame at the beginning of
atrial systole, Figure \ref{fig:pipeline}. A cut was made at the
ventricular base of the segmentation, so that the mesh cavity volume
and the ultrasound measured volume differed by less than  $1$ ml. Mesh
cells were marked into the 17 AHA regions through the regionally
delineated strain data, and the myocardial fiber orientation, denoted
by $\ef$, were assigned using the algorithm from Bayer et al \cite{bayer2012novel},
with the endo- and epicardial helix fiber angles set to
$\alpha_{\text{endo}} = 60$ and $\alpha_{\text{epi}} = -60$, respectively.

\subsection{Mechanical Model}
We represent the heart as a hyperelastic continuum body, where the coordinates in
the reference ($\Xvec$) and the current ($\xvec$) configuration are
related via the displacement field $\uvec = \xvec-\Xvec$. Furthermore,
we utilize the
deformation gradient, the determinant of the deformation gradient
and, the right Cauchy-Green deformation tensor given by $\F= \I+\nabla
\uvec $,  $J = \det \F$ and $\C = \F^T\F$, respectively. 
% Constitutive relations
To model the passive behavior of the myocardium, the transversely 
isotropic strain energy function proposed in  \cite{holzapfel2009constitutive} is adopted:
\begin{align}
\label{eq:hoa}
  \mathcal{W}(I_1, I_{4\ef}) = \frac{a}{2 b} \left( \exp\left\{ b (I_1 
  - 3)\right\}  -1 \right)
  + \frac{a_f}{2 b_f} \left( \exp\left\{ b_f (I_{4\ef} 
  - 1)_+^2\right\} -1 \right).
\end{align}
Here $I_1 = \tr \C$ and $I_{4\ef}= \ef \cdot (\C \ef)$ are invariants
of $\C$, $(\cdotp)_{+}  = \max\{\cdot, 0\}$, and $a, a_f, b, b_f$ are
material stiffness parameters defining the elastic properties of the myocardium.
% Incompressibility
We follow a common approach and assume that the myocardium is incompressible. 
Incompressibility is incorporated in the model by using a two-field variational 
approach, where we introduce a Lagrange multiplier $p$ which represents the 
hydrostatic pressure, and the term $p(J-1)$ is added to the strain-energy.

% Active response
To model the active response we apply the approach of active strain
\cite{ambrosi2012active}, which is based on decomposing
the deformation gradient into active and passive contributions, $\F =
\F_e \F_a$. We choose $\F_a = (1 - \gamma) \ef \otimes \ef +
\frac{1}{\sqrt{1 - \gamma}} (\I - \ef \otimes \ef)$, where $\gamma$
is a parameter that represents the relative active shortening along
the fibers. For reference, we have also performed tests with the
commonly used active stress formulation, where the stress tensor 
is additively decomposed into active and passive stress $\sigma =
\sigma_p + \sigma_a$. Here $\sigma_p$ is
the elastic material response, and $\sigma_a = T_a \Fef \otimes \Fef$ with
$\Fef = \F\ef$ and $T_a$ a scalar variable representing active fiber
tension.

For both approaches, the resulting displacement field $\uvec$ and hydrostatic pressure $p$
are determined by using the principle of stationary potential energy
\cite{holzapfel2000nonlinear}, which is based on minimizing the total
energy  $\Pi(\uvec, p)$, which includes internal energy derived from
\eqref{eq:hoa} and external energy. The external energy includes
contributions from the measured cavity pressure $p_{\text{LV}}$, and a
linear spring term at the basal boundary, with spring constant
$k$ = 10.0 kPa. The equilibrium solution is found by solving for the minimum
potential energy, $\delta\Pi(\uvec, p) = 0$.

@@

Here $V$ and $\varepsilon_{kj}$ are the measured volume and regional Lagrangian strain in
segment $j$ in direction $k$ respectively, and $\tilde{V}^i =
-\frac{1}{3} \int_{\partial \Omega_{\text{endo}}} (\Xvec + \uvec)
\cdotp J\F^{-T}\Nvec  \mathrm{d}S$,  and $\varepsilon_{kj} =
\frac{1}{|\Omega_j|}\int_{\Omega_j}  \mathbf{e}_k^T \nabla \uvec \,
\mathbf{e}_k  \mathrm{d}x.$ 
The parameters $\alpha$ and $\lambda$
control the weights on the different terms, and the sum in the second
term is taken over the seventeen AHA
segments, and the three different strain components (Section \ref{sec:clinical_data}).

The data assimilation procedure is divided into two phases; a passive
and an active phase. For the passive phase we iteratively
  estimate the unloaded configuration and the linear isotropic parameter, $a$ in
\eqref{eq:hoa}, using an algorithm similar to the one described in
\cite{nikou2016effects}, and we set $\alpha= 1.0$,  with $\lambda = 0$ and
$\gamma = 0$, minimizing only the misfit with the measured volumes. The
remaining material parameters are fixed according to [Table 1 row 3 of
\cite{holzapfel2009constitutive}]. For the active phase we fix the material parameter optimized in the
passive phase, choose the control variable $m$ to be $\gamma$ or
$T_a$ for the active strain and active stress model respectively, and
set $\alpha = 0.95$ and $\lambda = 0.01$. This choice of $\alpha$ and
$\lambda$ was based on the analysis done in \cite{balaban}. A summary
of our optimization pipeline is provided to the right in
Figure~\ref{fig:pipeline}.

\begin{figure}[htbp]
\centering
    \includegraphics[width=\textwidth]{models}
\caption{Left: Automated anatomical modeling pipeline to produce AHA
  marked simulation meshes with applied fiber orientations from 3D
  echocardiographic segmentations. Right: Optimization
  pipeline. 1. Unloaded geometry and the linear isotropic
  material parameter $a$ in \eqref{eq:hoa} are estimated iteratively. The unloaded geometry is
  estimated based on the backward displacement method (1a)
  \cite{nikou2016effects} and $a$ is estimated by minimizing the
  difference between simulated and measured volumes (1b). 2. The unloaded
  geometry and the material properties are fixed, and the amount of
  contraction ($\gamma$ for active strain and $T_a$ for active stress)
  is estimated by minimizing the mismatch between simulated and
  measured strain and volume. The active optimization continues to the
  next measurement point until all measurement points in the cycle are covered.}
\label{fig:pipeline}
\end{figure}

\subsection{Implementation details}
We employ a Galerkin finite element method with Taylor-Hood
tetrahedral elements, that is $(\uvec, p) \in \mathbb{P}_2 \times
\mathbb{P}_1 $, with $\mathbb{P}_n$ being the space of piecewise
polynomials of degree $n$.
The solver is implemented in the finite element framework FEniCS
\cite{logg2012automated}, and uses a Newton trust region algorithm
\cite{PETScPackage} to solve nonlinear systems. The minimization of the
model-data misfit functional~\eqref{eq:pde_opt} is accomplished by a sequential
quadratic programming algorithm (SQP)~\cite{kraft1988software}, where
the functional gradient is computed by solving an automatically derived adjoint
equation \cite{farrell2013automated}. In these minimizations an upper
bound of $0.5$ and $500$ kPa is set for the active strain ($\gamma$)
and active stress ($T_a$) control variable respectively, which both
are modeled as functions in $\mathbb{P}_1$, yielding one parameter per
nodal point in the mesh.

\subsection{Contraction analysis}
Although direct physical interpretation of the active
  strain parameter $\gamma$ is difficult, it may be seen as the
  relative shortening of an isolated and unloaded muscle cell. A high
  value of $\gamma$ is therefore an indication of higher contractile
  force in the myocardium, independent of load. We propose that
  the spatially averaged $\gamma$ over the entire LV, denoted by  
  $\overline{\gamma}$, can be used as an index of global
  contractility. Similarly, the active stress parameter $T_a$ is related to force
  development at level of the sarcomeres\cite{goktepe2014generalized},
  and the spatially averaged 
  $T_a$, denoted $\overline{T_a}$ can be used as an index of
  contractility. In addition to the contractility information contained in 
$\overline{\gamma}$ and $\overline{T_a}$, the overall elastic state
of the optimized patient models 
can be used to give estimates of LV elastance. The left ventricular
end-systolic elastance $E\es$, the response of end systolic volume to
increased load, is considered to be one of the major determinants for
cardiac systolic function, and was in \cite{sagawa1977end} proposed
as a global index of ventricular contractility.
It is possible to estimate the end systolic elastance directly 
if the end systolic pressure is known or estimated, 
by perturbing the loading conditions on the optimized model at end
systole while fixing the remaining quatities, and calculating the slope
in the resulting ES pressure-volume curve. More
precisely, if $\plv^{\text{ES}}$ is the end-systolic ventricular
pressure, with cavity volume $V^{\text{ES}}$, 
we change the pressure to $\plv^{\text{ES} + \Delta} =
\plv^{\text{ES}} + \Delta \plv$, resulting in a change in volume,
$V^{\text{ES} +\Delta} = V^{\text{ES}} + \Delta V$. The estimate of end-systolic
elastance can then be calculated by 
\begin{align}
  \tilde{E}\es = \frac{\Delta \plv}{\Delta V}.
  \label{eq:estimate_elastance}
\end{align}

@@

\subsection{Matching of strain and volume}
We show the results from two representative simulations in
Figure~\ref{fig:generic_result}, one from the LBBB group and one from
the healthy control group. Snapshots from the calculated unloaded and
end-systolic configurations are depicted. For the unloaded
geometry, we also show the image-based geometry at beginning of atrial
systole, and for the end-systolic configuration we show the
longitudinal strain using both the active stress and active strain
approach. We also show the agreement with the corresponding PV-loops.
% and one of the 51 strain traces over the cardiac cycle shown.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\textwidth]{generic_result}
  \caption{Left: Snap shots of the unloaded geometry in red, and the
    corresponding image based geometry taken at beginning of atrial
    systole in black wire-frame. Middle: Snap shot of end systolic
    configurations using the active strain and active stress approach.
    Color-map shows the end-systolic longitudinal strain.
    Right: Simulated and measured pressure-volume loops for
    these hearts using the active strain and active stress approach.}  
  \label{fig:generic_result}
\end{figure}

The total analysis of the 14 patients involved optimizing 432 volume
measurements and 20 853 strain measurements. The average time for one
forward and gradient evaluation was 8.3 and 8.9 seconds respectively
when running on a cluster using four cores, with an average number of
control parameters being 985.

In order to visualize the overall match of simulated to measured data,
we show linear regression plots in
Figure~\ref{fig:data_matching}. These results are all based on the
active strain formulation. For
the strain, we separately consider the diastolic and systolic points,
as different types of data were used to  constrain the model in these
two phases, namely volume in the diastolic phase and strain in the
systolic phase. An excellent overall fit was obtained for the optimized
volume ($R^2=1.00$) and systolic strains ($R^2=0.95$). 
Diastolic strains, not used in the optimization, were less well
matched ($R^2=0.31$).

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\textwidth]{data_matching}
  \caption{Scatter plot of simulated ($y$-axis) and measured
    ($x$-axis) strain and volume using the active
    strain approach. Left: Scatter plot of simulated versus
    measured volumes and the best linear least-squares fit of these
    points, (slope = $1.00$). Right: Scatter plot of simulated versus
    measured strain for all segments and all directions, separated
    into the diastole, were only the volume was optimized and systole,
    where both the strain and volume were optimized. For diastole, the slope of
    the best linear fit was $0.31$, while the best linear fit for the
    systolic points had a slope of $0.95$.}
  \label{fig:data_matching}
\end{figure}

@@

Global contraction time courses, $\overline{\gamma}$ and
  $\overline{T}_a$, for each patient were synchronized to the valvular
events to normalize for differing cycle lengths. The average 
and standard deviation of these normalized traces for the
LBBB vs the healthy controls are shown in left of
Figure~\ref{fig:contractility}. The healthy patients had a much higher
level of contraction through the cardiac cycle, and the peak values
were compared using one-way ANOVA, yielding a $P-$value less than
$0.001$ for both the active strain and the active stress approach.

The values of calculated $\tilde{E}\es$ for the healthy and LBBB patients are
shown to the right in Figure~\ref{fig:contractility}. The calculated
elastances of the LBBB group were significantly lower than for the healthy
group, with the comparison between the groups using one-way ANOVA
giving a $P-$value of $0.009$ and $0.003$ for the active strain and
active stress respectively.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\textwidth]{contractility}
  \caption{Extracted biomarkers related to cardiac contractility.
    Left:  Mean value of $T_a$ for the two groups
    synchronized with respect to valvular events (mvc: mitral valve
    closure, avo: aortic valve opening, avc: aortic valve closure,
    mvo: mitral valve opening). Shaded region shows $\pm$ one standard
    deviation.
    Middle:  Mean value of $\gamma$ for the two groups
    synchronized with respect to the same valvular events.
    Right: Estimated values of $\tilde{E}\es$, given by
    \eqref{eq:estimate_elastance} using the active stress and the
    active strain approach. The mean value is depicted for each
    group as a bar, and individual points are also displayed.} 
  \label{fig:contractility}
\end{figure}
